Intravital calcium imaging of meningeal macrophages reveals niche-specific dynamics and aberrant responses to brain hyperexcitability

Version 1

Abstract

The meninges, which envelop and protect the brain, host a dense network of resident macrophages with diverse roles in regulating homeostasis and neuroinflammation. Despite their importance, we have a limited understanding of their behavior in vivo. Many dynamic cellular functions of macrophages involve intracellular Ca2+ signaling. However, to our knowledge, virtually nothing is known about the spatiotemporal Ca2+ dynamics of meningeal macrophages. We developed a chronic intravital two-photon imaging approach and related computational analysis tools to interrogate meningeal macrophage Ca2+ dynamics, at a subcellular resolution, in a novel Pf4Cre:TIGRE2.0GCaMP6s/wt reporter mouse model. Using imaging in awake mice, we characterized the Ca2+ activity of meningeal macrophages at steady state and in response to cortical spreading depolarization (CSD), an aberrant pro-inflammatory brain hyperexcitability event, implicated in migraine, traumatic brain injury, and stroke. In homeostatic meninges, macrophages in the dural perivascular niche exhibited several Ca2+ dynamic features, including event duration and signal frequency spectrum, distinct from those of in the interstitial, non-perivascular niche. Simultaneous tracking of meningeal macrophage Ca2+ dynamics and local vasomotion revealed a subset of dural perivascular macrophages whose activity was coupled to behaviorally-driven diameter fluctuations of their associated vessels. Most perivascular and non-perivascular meningeal macrophages displayed propagating intracellular Ca2+ activity as well as synchronized intercellular Ca2+ elevations, likely driven by extrinsic factors. In response to CSD, the majority of perivascular and non-perivascular meningeal macrophages showed a persistent decrease in Ca2+ activity, while a smaller subset displayed Ca2+ elevations. Mechanistically, CGRP/RAMP1 signaling mediated the increase but not the decrease in CSD-mediated Ca2+ signaling. Collectively, our results highlight a previously unknown diversity of meningeal macrophage Ca2+ dynamics at steady state and in response to an aberrant brain hyperexcitability event linked to neuroinflammation.

Introduction

Resident macrophages are key myeloid immunocytes that play an important role in innate immune surveillance and defense across various peripheral tissues and organs [1–3]. The central nervous system also harbors a large subset of parenchymal macrophages, known as microglia, and several distinct subsets of macrophages localized to the brain’s border tissues, including the choroid plexus, perivascular spaces, and the meningeal compartments that cover, protect, and support the brain [4–8]. Macrophages are the predominant immune cell type within the brain meninges, and recent studies have demonstrated the diverse ontogeny, transcriptomic profiles, and immune functions of meningeal macrophages at steady state [5, 8–12] and in several neuropathological conditions [8, 13–16].

Cytoplasmic calcium (Ca2+) signaling underlies a wide variety of cellular homeostatic and inflammatory processes in macrophages [17–23]. In addition to intracellular Ca2+ elevation, distinct spatiotemporal dynamics - including oscillation patterns, intracellular propagations, and intercellular synchronization of Ca2+ signals - may regulate different macrophage functions during steady state and pathophysiology [20–25]. Despite our increased understanding of the diverse molecular signatures and contributions of meningeal macrophages to homeostasis and neuroinflammation, virtually nothing is known about their Ca2+ signaling heterogeneity in both healthy and diseased states.

To comprehensively characterize the Ca2+ dynamics of individual macrophages localized to the brain meninges, we combined intravital two-photon Ca2+ imaging in a novel reporter mouse line, in which the Ca2+ reporter GCaMP6s is expressed in platelet factor 4 (Pf4+) meningeal macrophages, with an event-based signaling analysis pipeline. Our data reveal several distinct spatiotemporal Ca2+ dynamic features in perivascular versus non-perivascular meningeal macrophages, including a unique coupling between the Ca2+ signals of dural perivascular macrophages and behaviorally-driven vasomotion of their associated dural vessels at steady state. Furthermore, our data uncover both increases and decreases in Ca2+ activity in distinct subsets of meningeal macrophages in response to cortical spreading depolarization (CSD), a pathophysiological brain hyperexcitability event linked to neuroinflammation [26] in migraine, traumatic brain injury, and stroke [27]. Mechanistically, our data suggest the calcitonin gene-related peptide/receptor activity-modifying protein 1 (CGRP/RAMP1) axis uniquely mediates CSD-evoked macrophage Ca2+ elevation, indicating a brain-to-meninges neuroimmune signaling pathway involving the activation of meningeal peptidergic sensory neurons.

Results

Characterizing macrophage Ca2+ signaling features in homeostatic brain meninges

Previous intravital imaging studies exploring the spatiotemporal dynamics of tissue-resident macrophages have primarily used CX3C motif chemokine receptor 1 (CX3CR1)-based mouse reporter strains [13, 22, 28, 29]. However, other resident monocyte-derived cells are labeled in these reporter mice [30, 31]. Moreover, brain microglia also express CX3CR1 [13, 31–33], limiting the use of these reporter mice for resolving the spatiotemporal subcellular Ca2+ dynamics of macrophages localized to the relatively thin meningeal layers covering the brain parenchyma. To systematically characterize meningeal macrophage Ca2+ dynamics in vivo and avoid contamination from Ca2+ signals arising from other resident monocyte-derived meningeal cells and superficial parenchymal microglia, we leveraged recent findings showing the expression of Pf4 in meningeal macrophages but not in parenchymal microglia, [10, 16, 34], and generated transgenic reporter mice expressing the highly sensitive Ca2+ indicator GCaMP6s (using TIGRE2.0-based Ai162D mice [35]) in Pf4+ macrophages (using Pf4Cre mice [36]). Notably, nearly 100% of meningeal macrophages are labelled in Pf4Cre-based reporter mice [34]. Although Pf4 is also expressed by megakaryocytes [36], these cells are absent in the brain meninges [5].

Anesthetic agents impact intracellular Ca2+ signaling, including in brain macrophages and other non-excitable glial cells [37–39]. We therefore investigated subcellular Ca2+ activity of meningeal macrophages in awake, behaving mice. We imaged meningeal macrophages via a chronic cranial window implanted together with a restraining headpost over the intact dura mater overlying the posterior neocortex. This chronic window approach produces minimal inflammatory responses in the cortex and meninges below the window [40, 41]. After at least 7 days of recovery, mice were gradually habituated across multiple days to head restraint while free to run on a wheel (Fig 1b). We tracked Ca2+ transients of GCaMP6s-labeled meningeal macrophages using high-speed two-photon microscopy in 37 fields of view (FOVs) from 7 mice. We first corrected the imaging movies for locomotion-evoked meningeal translational shifts using rigid registration [41]. Movies were then processed using the AQuA2 data analysis platform [42], which implements an unbiased event-based approach to capture spatiotemporal Ca2+ event dynamics (Fig 1C). Based on spatial analysis, we assigned events (n=1361), with their corresponding Ca2+ features, to a given macrophage, with data gathered from a total of 503 cells.

Imaging meningeal macrophage Ca2+ dynamics in awake behaving mice.

(A) Pf4Cre:GCaMP6sfl/wt reporter mouse construct for imaging meningeal macrophages Ca2+ activity. (B) Experimental procedure for two-photon imaging of meningeal macrophage Ca2+ activity. Following the implantation of a headpost and a cranial window, mice were habituated to head restraint and subjected to two-photon microscopy while head-fixed on a running wheel to study meningeal macrophage Ca2+ activity. (C) Macrophage Ca2+ imaging processing pipeline.

We characterized Ca2+ dynamics in two anatomically distinct macrophage populations: perivascular and interstitial non-perivascular (n=122; n=381, respectively, Fig 2A and Video S1). The vast majority of perivascular macrophages (93.4%, n=114) displaying ongoing Ca2+ activity were associated with vessels in the dura mater. The two meningeal macrophage subpopulations exhibited several distinct Ca2+ activity features. While the total area of Ca2+ activity in the peri- and non-perivascular macrophages was similar (Fig 2B), the signal perimeter of perivascular macrophages was significantly higher (Fig 2C) and exhibited a more elongated shape (Fig 2A, 2D). The Ca2+ event duration in the perivascular macrophages was also longer compared to the non-perivascular subpopulation (Fig 2F). The peak Ca2+ activity level (Max DF/F0, Fig 2E) and event rate (Fig 2G), nonetheless, were similar in the two meningeal macrophage subpopulations.

Ca2+ dynamic features of meningeal macrophages at steady state.

(A) Left: Mean projection of an example FOV depicting perivascular (P, red; 1, 2) and non-perivascular (NP, blue; 3, 4) meningeal macrophages (white squares). Scale bar 50 μm. Right: Corresponding macrophages with representative 900-second Ca2+ activity traces (top) and their fluorescence signal at selected time points (bottom). (B-G) AQuA2-based morphological and Ca2+ event functional features of perivascular (P=122 cells) and non-perivascular (NP, n=381 cells) meningeal macrophage. (B) Event area, (C) Event perimeter, (D) Event circularity, (E) Event max DF/F0, (F) Event duration, (G) Event rate. Data (B-G) represent median ± IQR. ***p<0.001, ****p<0.0001, Mann-Whitney U-test. n=37 fields of view (FOVs) from 7 mice.

Distinct Ca2+ signal frequency spectra may underlie different biological functions of macrophages [22]. We therefore analyzed meningeal macrophage Ca2+ signal waveforms by performing a multi-step signal processing and clustering analysis. We clustered cells based on two distinct patterns of Ca2+ activity. Cells in Cluster 1 (n=40) exhibited a more noisy-like activity pattern characterized by multiple frequencies, while cells in Cluster 2 (n=463) displayed a single dominant frequency at 0.01 Hz (Fig 3A). To further explore these cellular signaling differences, we combined the clustering data with AQuA2-derived features and observed that macrophages in Cluster 1 showed a larger event perimeter and lower circularity and peak magnitude compared to those in Cluster 2 (Fig 3D-F). Cluster 1 cells also had a lower signal-to-noise ratio compared to Cluster 2 cells (Fig 3I), but had differences in terms of event area, duration, or rate (Figures 3C, G, H). There was a significant association between cell cluster (1 vs. 2) and cell type, with Cluster 1 predominantly comprised of perivascular macrophages, and Cluster 2 including primarily non-perivascular macrophages (Fig 3J), further indicating that these two meningeal macrophage subpopulations have distinct Ca2+ signaling properties.

Intracellular Ca2+ signal frequency spectra of macrophage subsets in the steady-state meninges

(A) Left: Clustering of meningeal macrophage Ca2+ activity based on frequency-domain features and peak detection. Cluster 1 (purple, n=40) and Cluster 2 (green, n=463). Right: Power spectrum density (PSD) of Ca2+ signals for each cluster. (B) Example DF/F0 heatmaps of Cluster 1 and Cluster 2 macrophages. (C-H) Morphological and Ca2+ functional features of Cluster 1 and Cluster 2 macrophages. (C) Event area. (D) Event perimeter (E) Event circularity. (F) Event max DF/F0. (G) Event duration. (H) Event rate. (I) Signal-to-noise ratio from Clusters 1 and 2. (J) Distribution of cell types across clusters. Data (C-I) represents median ± IQR. ***p<0.001, ****p < 0.0001, Mann-Whitney U-test Data (J) represents the cell proportion. ****p<0.0001, Chi-square test.

Intracellular Ca2+ signal propagation underlies diverse cellular functions and has been recently identified in macrophages in vitro [21] and in skin-resident macrophages in vivo [43]. By assessing the Ca2+ signal propagation maps of each event within a determined cell, we identified two distinct patterns of activity: propagating events in which Ca2+ signals traveled throughout the entire cell, and stationary events (Fig 4A). Propagating events had varied signal source regions and directionality. Most macrophages (perivascular, 94.3%, n=115; non-perivascular, 86.9%, n=331) exhibited only propagating events, while a small minority of cells displayed a mix of propagating and stationary events or exclusively stationary activity events (Fig 4B).

Intracellular propagation and intercellular synchronization of meningeal macrophage Ca2+ activity.

(A) Spatial maps of two distinct Ca2+ events. Left: Propagating Ca2+ activity. Right: Stationary Ca2+ activity. (B) Distribution of event propagation profiles in perivascular and non-perivascular meningeal macrophages. (C) Schematic analysis paradigm for detecting synchronous Ca2+ activity in meningeal macrophages. (D) Synchronous Ca2+ events among meningeal macrophages within a FOV. Left: Mean projection of an example FOV showing Ca2+ activity in distinct macrophages (colored/numbered). Scale bar 50 μm. Right: Spatial map of macrophages exhibiting synchronous Ca2+ activity. Lines connect macrophages with synchronized Ca2+ activity, and colors indicate the extent of Ca2+ event synchronization. (E) Distribution of distances across macrophage pairs showing different numbers of synchronous Ca2+ events. (F) Linear regression showing poor correlation between distances of macrophage with synchronized Ca2+ activity and event delay. (G) Proportion of macrophages exhibiting a specific synchronized interaction (cells interacting only with the same subtype: P-P/NP-NP; cells interacting only with a different subtype: P-NP; mixed (cells interacting with the same and different subtypes).

Macrophage intercellular communication, including synchronized activity, potentially across connected macrophage networks, has been implicated in maintaining tissue homeostasis and immune function [18, 21, 44, 45]. To investigate meningeal macrophage intercellular interactions, we characterized spatiotemporal relationships between Ca2+ events in distinct cells within each FOV. We compared temporal factors, including the relative onset latency (t0A – t0B) between different macrophages and the duration of the first occurring event (t1A – t0A) (Fig 4C). We also compared the distances between macrophage pairs exhibiting concurrent events and the number of synchronous events (Fig 4D). Finally, we calculated the proportion of perivascular and non-perivascular macrophages exhibiting synchronous Ca2+ events. Across all FOVs, nearly half of the macrophages (49.3%) exhibited synchronous Ca2+ activity, involving cells at a wide range of distances (Fig 4E), suggesting that spatial proximity does not influence event synchronicity. Interestingly, there was poor correlation between the distance of cell pairs exhibiting synchronous events and their timing (Fig 4F), suggesting that an external factor, rather than intercellular coordinated networked activity, mediates this intercellular signaling. When comparing synchronous activity among peri- and non-perivascular macrophages, most cells exhibited mixed interactions between the same and different subtypes (Fig 4G).

Dural perivascular macrophage Ca2+ activity is tuned to behaviorally-driven dural vasomotion

Brain border macrophages in the leptomeninges and related parenchymal perivascular spaces regulate pial arterial vasomotion indirectly by affecting vessel stiffness [6]. Yet, interaction between vascular-associated macrophages residing in the dura mater, the outermost meningeal layer, and dural vasomotion remains unknown. We employed a Generalized Linear Model (GLM) approach to investigate functional interaction between dural perivascular macrophage Ca2+ signals and behaviorally-driven dural vessel dynamics (Fig 5A). Dural arteries constrict during locomotion, while pial arteries dilate [46]. We analyzed the locomotion-associated responses of 86 meningeal vessels (32 FOVs from 5 mice) and identified a subset (22%, n=19), whereby the diameter changes were well fit by a GLM using the locomotion state as a predictor. Of these, we identified 74% (n=14) as dural vessels based on their GLM’s negative coefficients consistent with constriction (Fig 5B and 5C). Next, we fitted the Ca2+ signal observed in perivascular macrophages associated with these dural vessels (n=35) to a GLM using the diameter changes as a predictor variable. Overall, the Ca2+ activity of 83% (n=29) of these dural macrophages was well predicted by the model (average deviance explained across all well-fit macrophages: 0.43±0.15, mean±SD). Analysis of the macrophage-vascular models’ beta coefficients revealed two distinct interactions. About half of the macrophages (55%, n=16) exhibited negative coefficients (i.e., increase and decrease in Ca2+ activity associated with dural vasoconstriction and recovery, respectively; Fig 5B and 5D). The remaining macrophages (45%, n=13) exhibited positive coefficients (i.e., decrease and increase in Ca2+ activity in response to dural constriction and recovery, respectively; Fig 5B and 5 E). The coefficients for increased and decreased macrophage Ca2+ activity peaked near zero delay relative to the vasoconstriction and were not statistically different (Fig 5D-E and 5H). These data provide evidence that dural perivascular macrophages are functionally coupled to dural vasomotion, either responding to or mediating such behaviorally driven vascular response.

Ca2+ signals of dural perivascular macrophages are functionally coupled to behaviorally driven dural vasomotion.

(A) Experimental paradigm: Locomotion data were acquired during imaging in awake-behaving mice. Behaviorally-evoked changes in meningeal vessel diameter were obtained using segmentation of vessels labeled with a tracer during macrophage Ca2+ and further tested for coupling with Ca2+ signals. (B) Example data of macrophages with Ca2+ activity tuned to locomotion-related dural vessel vasomotion. Locomotion bouts (top trace) and dural vessel diameter (black trace) that was well fit by a GLM (red line) using locomotion state as a predictor. Note vasoconstriction during locomotion, indicating a dural vessel. The two bottom traces depict the Ca2+ signal (black traces) of distinct meningeal macrophages that were well-fit by a GLM using dural vessel diameter as a predictor, showing either a negative coefficient (purple) or a positive coefficient (green). (C) Temporal profile of dura vessels GLM coefficients averaged across all well-fitted vessels (n=19). Traces represent the median across all well-fit ROIs, with shaded regions indicating IQR. (D, E) Temporal profiles of GLM coefficient values for macrophages’ Ca2+ activity averaged across all well-fitted cells (purple, negative coefficient, n=16; green, positive coefficient, n=13). Traces represent the median across all well-fit macrophages, with shaded regions indicating IQR. (F) The deviance explained (goodness-of-fit estimates) of the vessel diameter data included in the GLM used to predict macrophage Ca2+ activity were not statistically different for macrophages showing negative (n=16) and positive (n=13) coefficients, suggesting similar interaction levels. (G) Comparison of peak positive and negative coefficients of well-fit macrophage Ca2+ activity/vasomotion GLMs. (H) Center of mass of GLM coefficients indicating that dural vessel diameter changes drive bidirectional changes in fluorescence signal at zero delay. Data (F-H) are median ± IQR. ****p<0.0001 (Mann-Whitney U-test).

An acute aberrant pro-inflammatory brain hyperexcitability event drives diverse Ca2+ dynamics in meningeal macrophages

CSD is an aberrant brain hyperexcitability event linked to parenchymal inflammation and pain in migraine, traumatic brain injury, and stroke [27, 47, 48], and could also affect meningeal macrophages [49]. In anesthetized mice subjected to an acute CSD episode, a small subset of meningeal macrophages undergoes morphological changes resembling an inflammatory state [26]. Given the direct anatomical and functional connection between the brain and meninges [11, 50] and the involvement of intracellular Ca2+ influx in macrophage inflammatory activation [51], we asked whether brain-to-meninges signaling drives intracellular Ca2+ elevations in meningeal macrophages in the wake of CSD. We therefore triggered a single CSD episode in the frontal cortex [52, 53] and characterized the related changes in meningeal macrophage Ca2+ dynamics. In each experiment, we verified CSD induction based on the associated acute meningeal deformation and/or pial vasoconstriction observed in awake animals ([53] and Video S2). We studied CSD-related Ca2+ dynamics in 249 macrophages (perivascular, n=64; non-perivascular, n=185; 13 FOVs from 10 mice). For each cell, we compared Ca2+ event rates during the passage of the CSD wave (1 min) and the PostCSD period (30 min) to baseline (PreCSD, 30 min) to assess acute and persistent Ca2+ responses, respectively. We characterized cells as exhibiting persistently increased (event rate > 2× PreCSD), decreased (event rate < 0.5× PreCSD), or unchanged responses. Given the low Ca2+ activity rate observed under steady state and the likelihood that no spontaneous Ca2+ elevations occur during the brief period of the CSD event, we considered cells to be either acutely activated or exhibiting an unchanged Ca2+ response. CSD evoked both acute and persistent Ca2+ activity changes in (Fig 6 A-E and Video S2). While smaller subsets of meningeal macrophages exhibited acute (21.3%, n=53) and/or persistent increase (22.1%, n=55) in their Ca2+ activity, the majority of cells (58.6%, n=146) exhibited a persistent decrease. An acute increase was observed more often in peri-vs non-perivascular macrophages (perivascular, 32.8%, n=21; non-perivascular, 17.3%, n=32, Fig 6F). Persistent changes in Ca2+ activity were observed similarly in peri- and non-perivascular macrophages (increases; perivascular, 28.2%, n=18; non-perivascular, 18.4%, n=37; decreases; perivascular, 50.0%, n=32; non-perivascular, 61.6%, n=114, Fig 6G). The macrophages’ propensity to develop a persistent Ca2+ increase was unrelated to their acute response (Fig 6H), suggesting that the mechanisms underlying these two temporal responses are distinct. However, cells showing no acute activation were more likely to display decreased Ca2+ activity post-CSD (Fig 6H). Finally, we observed that macrophages exhibiting a persistent increase in Ca2+ activity had a lower baseline activity when compared to cells showing a persistent (Fig 6I), suggesting that this post-CSD response is influenced by the macrophages’ basal response activity.

Diverse meningeal macrophage Ca2+ dynamics following cortical spreading depression (CSD).

(A) Experimental setup and example data: (top) A small burr hole was drilled above the frontal cortex 7 days before Ca2+ imaging. Mice were pretreated with saline 30 min before imaging baseline macrophage Ca2+ activity (30 min, PreCSD). CSD was then induced with a pin prick, and macrophage Ca2+ activity was assessed during CSD (1 min during CSD), and post-CSD (30 min, post-CSD). (Bottom) A raster plot of macrophage Ca2+ activity showing the acute and persistent increases and persistent decrease in response to CSD. (B) Example of macrophage Ca2+ fluorescence changes following CSD. Images depict the mean projection over the specific experimental timeline. Arrows indicate an increase or a decrease in Ca2+ activity. Scale bar 50 μm. (C) Individual responses of perivascular (P, n=21) and non-perivascular (NP, n=32) macrophages showing an acute increase in Ca2+ activity. (D) Individual response of P (n=18) and NP (n=37) macrophages exhibiting a persistent increase in Ca2+ activity. (E) Individual responses of P (n=32) and NP (n=114) macrophages showing a persistent decrease in Ca2+ activity. (F) Proportion of P and NP macrophages showing an acute increase in Ca2+ activity or no acute change. (G) Proportion of P and NP macrophages showing a persistent increase, decrease, or no change in Ca2+ activity. (H) Proportion of macrophages displaying distinct persistent responses stratified based on their acute response. (I) Baseline (PreCSD) Ca2+ activity in macrophages exhibiting persistent increased (n=55) or decreased activity (n=146). Data (C-E) are median ± IQR. ****p<0.0001 (Wilcoxon signed rank test). Data (F-H) represents the proportion of cells, **p<0.01; ****p<0.0001 (Chi-square test). Data (I) are median ± IQR. ****p<0.0001 (Mann-Whitney U-test).

CGRP/RAMP1 signaling mediates CSD-evoked persistent increase in meningeal macrophage Ca2+ activity

Many meningeal macrophages are localized near peptidergic sensory axons [16]. In the wake of CSD, cortex-to-meninges signaling enhances the responsiveness of meningeal sensory neurons and can drive the release of the neuropeptide CGRP from their peripheral nerve endings [52–54]. CGRP-expressing sensory neurons regulate tissue immunity, including meningeal macrophage function, via the CGRP/RAMP1 neuroimmune axis [16, 55]. We therefore asked whether the CSD-related changes in meningeal macrophage Ca2+ dynamics we observed involve CGRP/RAMP1 signaling. We pretreated mice with the RAMP1 antagonist BIBN4096 and then imaged meningeal macrophage Ca2+ activity (42 cells; perivascular, n=14; non-perivascular, n=28; 3 FOVs from 3 mice) before and after CSD. As expected, RAMP1 blockade did not affect CSD triggering [56]. When compared to control saline treatment, RAMP1 blockade also did not reduce basal macrophage Ca2+ activity (Fig 7B). Blocking CGRP/RAMP1 signaling neither affected the incidence of acute increases in Ca2+ activity (Fig 7A and 7C) nor the magnitude of that response (Fig 7D). RAMP1 antagonism, however, inhibited the CSD-evoked persistent increase in the macrophage’s Ca2+ activity, without affecting the incidence of the persistent decrease (Fig 7E). The data suggest that in the wake of CSD, the CGRP/RAMP1 axis, likely involving a peripheral sensory neuroimmune signaling, is responsible for the prolonged enhancement of Ca2+ signaling in a subset of meningeal macrophages, which could potentially mediate their pro-inflammatory response.

CGRP/RAMP1 signaling mediates CSD-related persistent increase in meningeal macrophage Ca2+ activity.

(A) Experimental setup and example data: (top) A small burr hole was drilled above the frontal cortex 7 days before Ca2+ imaging. Mice were pretreated with the RAMP1 antagonist BIBN4096 (BIBN) 30 min before imaging baseline macrophage Ca2+ activity (30 min, PreCSD). CSD was then induced with a pin prick, and macrophage Ca2+ activity was assessed during CSD (1 min during CSD), and post-CSD (30 min, post-CSD). (Bottom) A raster plot of macrophage Ca2+ activity showing the acute and persistent increases and persistent decrease in response to CSD. (B) RAMP1 inhibition does not affect the baseline (PreCSD) event rate. Data compared between macrophages imaged in saline-treated mice (n=249 cells) and BIBN-treated mice (n=42 cells). (C) RAMP1 antagonism does not affect the CSD-evoked acute increase in macrophage Ca2+. Proportion of macrophages showing an acute response (increase vs. no change) in saline-treated mice (n=249 cells) and BIBN-treated mice (n=42 cells). (D) RAMP1 antagonism does not affect the magnitude of the acute macrophage Ca2+ signal. Event max DF/F0 in macrophages showing an acute Ca2+ increase in saline-treated mice (n=53 cells) and BIBN-treated mice (n=8 cells). (E) RAMP1 antagonism distinctly inhibits the persistent increase in macrophage Ca2+ activity post CSD. Proportion of macrophages showing persistent increase, persistent decrease, or no persistent change in saline-treated mice (n=249 cells) and BIBN-treated mice (n=42 cells). (F) Data (B, C, E) represents the proportion of cells. *p<0.05, Fisher’s exact test. Data (D) are median ± IQR.

Discussion

Resident macrophages in the brain meninges are essential for maintaining brain homeostasis, regulating central nervous system immune surveillance, and mediating neuroimmune responses under pathological conditions [12, 14, 16, 57–59]. Macrophages rely on Ca2+ signaling to mediate many of their functions [17–23]. Here, using two-photon microscopy in awake, behaving Pf4Cre:GCaMP6s reporter mice, we describe the heterogeneity of meningeal macrophage Ca2+ dynamics at steady state and in response to CSD, an aberrant cortical hyperexcitability event associated with migraine, traumatic brain injury, and stroke. Our data suggest the presence of two subsets of macrophages in discrete meningeal niches, perivascular and non-perivascular, that exhibit several distinct Ca2+ signal properties at steady state. We further demonstrate Ca2+ activity in dural perivascular macrophages tuned to behaviorally-driven dural vasomotion. Finally, we describe opposing Ca2+ responses of meningeal macrophages following CSD, and demonstrate the contribution of the CGRP/RAMP1 axis in mediating CSD-evoked persistent Ca2+ elevations.

The exact link between the distinct Ca2+ signal properties of meningeal macrophage subsets observed herein and their homeostatic function remains to be established. The lower event magnitude and noisier signal observed in dural perivascular macrophages may reflect functional interactions with the pulsation dynamics of dural vessels [46]. Indeed, by combining vascular and macrophage Ca2+ imaging, we demonstrate a tight temporal association between the diameter fluctuations of the dural vessels and the Ca2+ signal of their associated macrophages in awake, locomoting mice. Direct vascular-macrophage coupling may underlie this interaction, involving macrophages sensing vascular-related mechanical changes via Piezo1 signaling [60]. Functional vascular-macrophage interaction may also involve mural cells as inter-mediators [58, 61]. Studying whether macrophage Ca2+ signaling regulates dural vasomotion will require an experimental approach that has yet to be developed, allowing for their selective manipulation. The paucity of ongoing Ca2+ activity in perivascular macrophages situated in the leptomeninges we observed supports recent findings that subdural perivascular macrophages indirectly affect pial and parenchymal vasomotion via extracellular matrix remodeling [6].

Intracellular Ca2+ signal propagation has been observed in various non-excitable cells, such as astrocytes [62]. We show that the majority of meningeal macrophages, including peri- and non-perivascular, exhibit intracellular Ca2+ signals that propagate throughout the entire cell, suggesting microdomain intracellular Ca2+ signal elevation, following release from internal stores. By characterizing the spatiotemporal relationships between Ca2+ signals in distinct cells, we also demonstrate synchronous events that are independent of spatial proximity, suggesting that synchronous Ca2+ activity is not driven by intercellular communication but rather by an external signal. Interestingly, our data indicate that synchronized events involve both peri- and non-perivascular macrophages, despite having distinct Ca2+ elevation features, suggesting that these meningeal macrophage subtypes similarly sense external signals.

Cortex-to-meninges signaling involves a relatively slow flow of soluble molecules in the cerebrospinal fluid reaching the subarachnoid space and advancing via arachnoid cuff exit points into the dura mater [11, 50]. Our finding of an acute Ca2+ elevation in a subset of extrasinusoidal perivascular dural macrophages coinciding with the CSD event suggests a rapid transfer of soluble signaling factors from the cortex across all meningeal layers [47]. Nevertheless, we cannot exclude a mechanically driven macrophage response to the acute meningeal deformation produced by the neuronal and glial swelling and shrinkage of the cortical extracellular space during CSD [53, 63, 64]. Our data also indicate a delayed and prolonged increase in Ca2+ signal in a relatively small subset of macrophages post-CSD, which could underlie their proinflammatory-like morphological change [49, 51]. Our findings also support the view that meningeal neuroimmune CGRP/RAMP1 axis serves as a mechanism responsible for this macrophage Ca2+ response, likely via the activation of their CGRP receptors [16]. Whether the relatively small subset of meningeal macrophages featuring increased Ca2+ signaling serves a protective role [65, 66] or a proinflammatory, destructive function [67] remains to be elucidated. Intriguingly, our data points to a persistent decrease in macrophage Ca2+ activity post-CSD, not involving CGRP/RAMP1 signaling, as the most prevalent response, raising the possibility that a hyperexcitable cortex dampens meningeal immunity.

Conclusions

We provide a detailed characterization of macrophage Ca2+ dynamics in homeostatic meninges, thereby expanding our understanding of their biological diversity. The coupling of dural perivascular macrophage Ca2+ signals and dural vasomotion may represent a unique homeostatic functional dural macrophage-vascular unit that controls dural perfusion. The diversity of meningeal macrophage Ca2+ responses to CSD further highlights the complexity of brain-to-meninges neuroimmune signaling and meningeal macrophage function in neurological disorders such as migraine, traumatic brain injury, and stroke. Our study also provides essential genetic and data analysis tools to further understand the molecular signaling underlying macrophage function at steady state and neuropathological conditions.

Methods

Animals

All experimental procedures were approved by the Beth Israel Deaconess Medical Center Institutional Animal Care and Use Committee. Experiments were conducted on adult Pf4Cre/wt:TIGRE2.0GCaMP6s/wt Ca2+ reporter mice (8-17 weeks) (9 males, 5 females). Mice were generated by crossing Pf4Cre (C57BL/6-Tg (Pf4-icre) Q3Rsko/J, Jackson laboratory, Strain #008535) mice with GCaMP6sfl/fl (Ai162D; B6.Cg-Igs7tm162.1(tetO-GCaMP6s,CAG-tTA2)Hze/J, Jackson laboratory, Strain #031562) mice. Animals were genotyped by Transnetyx Inc.

Surgical procedures

Animals were anesthetized using isoflurane in 100% O2 (induction: 3%; maintenance: 1.5-2%) and placed on a heating pad with a rectal probe attached to a stereotaxic frame to monitor animal body temperature during surgery. Animals received dexamethasone (8 mg/kg, i.p.) and Meloxicam SR (4 mg/kg, s.c.) to reduce inflammation and improve surgical outcomes. An eye ointment was used to prevent ocular drying. Mice were implanted with a titanium headpost and a 3 mm glass cranial window (1.5 mm lateral and 2 mm posterior to Bregma) over an intact dura covering the left posterior neocortex [53]. Immediately after surgery, the mouse cage was placed on a water-circulating heating pad for faster recovery. Animals were then single-housed with access to a running wheel and a hut and allowed to recover for at least one week.

Wheel running acclimation

After the cranial window surgery, mice were allowed to recover for at least a week. To reduce stress associated with head-fixation during imaging and habituate to wheel running, the mice received multiple training sessions (10 min to 1 h over 3–4 days). In each session, the mouse was placed on a 3D printed running wheel, with its headpost attached to two clamps, and allowed to locomote freely.

Two-photon Ca2+ imaging

Awake-behaving mice were head-fixed to the running wheel by its headpost (Figure 1). We used a two-photon microscope (Neurolabware) with a Nikon 16X, 0.8 NA objective to acquire images at 15.5 Hz with digital zoom set at 4X (312 x 212 µm2 FOV). A MaiTai laser set to 920 nm with 25–40 mW power was used to excite fluorescence. The Scanbox package for MATLAB (Neurolabware) was used to control the microscope and acquire images and wheel running data. To image the meningeal vasculature, mice were administered 70kDa TRITC-Dextran tracer (50 mg/kg, i.v.; Sigma-Aldrich).

Behavioral tracking during two-photon imaging

We recorded running speed in MATLAB using a custom-made encoder (Arduino) coupled to the 3D-printed running wheel. See below for details of analyses of behavioral variables.

Induction of CSD and pharmacological treatment

For CSD induction, a 1 mm burr hole was drilled at the frontal bone (1.5 mm anterior to the cranial window) to allow access to the brain cortical surface. A small amount of silicone elastomer (Kwik-Cast, WPI) was placed to cover the burr hole opening, and the animal was left to recover for at least one week. CSD was induced using a brief 2-second cortical pinprick [53]. CSD induction was confirmed by the identification of a short-lasting meningeal deformation and/or transient pial constriction [53]. On the experimental day, 30 minutes before baseline recording, mice were pretreated with the CGRP/RAMP1 antagonist BIBN4096 (0.3 ml, 1 mg/kg, i.p., Tocris) [54] or 0.3 ml of saline (Vehicle control).

Quantification and statistical analysis

Two-photon imaging movie processing

We used a discrete Fourier transform to perform rigid registration to correct translation changes due to brain motion during locomotion. Movies were then downsampled to 1.03 Hz. Locomotion signals were detected as described [41]. All image processing and locomotion signal extraction were performed in MATLAB 2021b (Mathworks).

Ca2+ signal detection pipeline

For detecting macrophage Ca2+ signals, we used the Activity Quantification and Analysis (AQuA2) platform that implements an event-based approach with advanced machine learning techniques for temporal and spatial segmentation of Ca2+ fluorescence events38. Importantly, this computational platform captures event dynamics beyond traditional ROI-based approaches. The following user-defined parameters were input: 0.49 µm/pixel spatial resolution, 1.03 Hz temporal resolution, 1-second minimal event duration detection; window of event size between 25% (125 µm2/ 523 pixels) and 150% (750 µm2/ 3140 pixels). Every event and cell identified was followed by a manual visual check. Subsequently, each cell was labelled as perivascular or non-perivascular according to its location relative to vessels. Morphological features (area, perimeter, circularity) and spatiotemporal aspects of the Ca2+ signals (i.e., ι1F/F0 dynamics, frequency, amplitude, duration) were used to analyze cell-specific characteristics and Ca2+ activity profiles. We employed the AQuA2 automatically-generated function [42] to characterize intracellular Ca2+ propagation. Intercellular Ca2+ activity was evaluated by analyzing temporally co-occurring events (synchronized event pairs), as well as their corresponding spatial localization in the FOV. All post-processing of AQuA2-generated data was performed using MATLAB 2021b (Mathworks).

Clustering of Ca2+ dynamics

We used a Savitzky-Golay filter to detrend and smooth Ca2+ activity traces. Polynomial order and frame size were optimized for each cell by selecting the combination of parameters that yielded a better signal-to-noise ratio. From the filtered signals, we extracted the dominant frequency using Fast Fourier Transform, and peak counts using minimum peak prominence (threshold of 10% signal amplitude range). The optimal number of clusters was determined using the Elbow method, after which k-means clustering was applied to group cells based on their signal characteristics.

Analysis of CSD-related changes in macrophage Ca2+ dynamics

For analyzing the effects of CSD on macrophages’ Ca2+ dynamics, we divided the activity of each cell into three phases: ‘PreCSD’ (minutes 0 – 30), ‘DuringCSD’ (minutes 30 – 31), and ‘PostCSD’ (minutes 31 – 61). Ca2+ event rate during CSD and PostCSD was compared against that of the PreCSD baseline for evaluation of acute and persistent Ca2+ responses, respectively. Ca2+ responses were categorized as increased (event rate > 2x PreCSD), decreased (event rate < 0.5x PreCSD), or unchanged.

Analysis of locomotion

We extracted running speed (cm/sec) from the wheel encoder. To infer the locomotion state, we first concatenated all velocity signals obtained from a given mouse across all experiments and trained a two-state Hidden Markov Model using the MATLAB function ‘hmmtrain’. Then, the locomotion state was inferred for each individual imaging run by applying the MATLAB function ‘hmmviterbi’ with the model trained on the concatenated data. Locomotion bouts were defined as periods when the locomotion state was sustained for at least two seconds.

Vascular signals

We calculated changes in vascular diameter by first generating a maximum intensity projection of the red channel and drawing polygons (ROIs) around each vessel. Subsequently, for each frame, pixels inside each ROI were extracted, and a Radon transform was applied to get a 1D vessel profile. Radon transform of the first frame was used as a reference to normalize the results of all subsequent frames, resulting in a time series of normalized diameter traces.

General linear models (GLM)

We investigate the functional interaction between dural perivascular macrophage Ca2+ activity and dural vasomotion by fitting a Gaussian GLMs using the GLMnet package (MATLAB 2021b), with elastic net regularization (α=0.01) and ten-fold cross-validation. We employed a two-step modeling. The first model was used to identify dural vessels by evaluating the correlation of vessel diameter changes to locomotion state, classifying dural or pial vessels according to their vasoconstriction and vasodilation dynamics, respectively [46]. The second modeling step evaluated the relative contribution of perivascular dural Ca2+ activity signals to dural vessel diameter changes. All signals were downsampled to 1.03 Hz to match the frame rate used when extracting fluorescence Ca2+ signals in AQuA2. To allow potential anticipatory or delayed responses of diameter- to-locomotion or Ca2+ fluorescence-to-diameter, the time interval used to analyze the response to the predictor was set from -60 s to +60 s. The GLM was trained on 75% of the data, and all predictions and model performances reported are from the remaining 25% testing set. A threshold of 0.1 goodness-of-fit deviance explained was set. For each predictor temporal shift, a response coefficient was generated. Centroid delay between predictor and response was calculated, weighted by the absolute value of the coefficient (i.e., center-of-mass, COM).

Statistical analysis

All statistical analysis was performed using GraphPad Prism 10.4 and MATLAB 2021b. Data were analyzed using a Wilcoxon matched-pairs signed rank sum test or a Mann-Whitney U-test. Distribution of categorical data was analyzed using the Chi-square or Fisher’s exact test. P-values are indicated as follows: *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001.

eLife Assessment

This valuable study presents a technically sophisticated intravital two-photon calcium imaging approach to characterize Ca²⁺ dynamics in distinct populations of meningeal macrophages in awake, freely behaving mice. These data are solid and suggest that meningeal macrophage calcium activity is tightly linked to anatomical sub-compartments, with potential implications for migraine and neuroinflammatory processes. Despite these strengths and broad relevance to neuroimmunology, several technical and interpretational issues limit the study, which could be addressed to strengthen this manuscript.

Reviewer #1 (Public review):

Summary:

This study presents a technically sophisticated intravital two-photon calcium imaging approach to characterize meningeal macrophage Ca²⁺ dynamics in awake mice. The development of a Pf4Cre:GCaMP6s reporter line and the integration of event-based Ca²⁺ analysis represent clear methodological strengths. The findings reveal niche-specific Ca²⁺ signaling patterns and heterogeneous macrophage responses to cortical spreading depolarization (CSD), with potential relevance to migraine and neuroinflammatory conditions. Despite these strengths, several conceptual, technical, and interpretational issues limit the impact and mechanistic depth of the study. Addressing the points below would substantially strengthen the manuscript.

Strengths:

The use of chronic two-photon Ca²⁺ imaging in awake, behaving mice represents a major technical strength, minimizing confounds introduced by anesthesia. The development of a Pf4Cre:GCaMP6s reporter line, combined with high-resolution intravital imaging, enables long-term and subcellular analysis of macrophage Ca²⁺ dynamics in the meninges.

The comparison between perivascular and non-perivascular macrophages reveals clear niche-dependent differences in Ca²⁺ signaling properties. The identification of macrophage Ca²⁺ activity temporally coupled to dural vasomotion is particularly intriguing and highlights a potential macrophage-vascular functional unit in the dura.

By linking macrophage Ca²⁺ responses to CSD and implicating CGRP/RAMP1 signaling in a subset of these responses, the study connects meningeal macrophage activity to clinically relevant neuroimmune pathways involved in migraine and other neurological disorders.

Weaknesses:

The manuscript relies heavily on Pf4Cre-driven GCaMP6s expression to selectively image meningeal macrophages. Although prior studies are cited to support Pf4 specificity, Pf4 is not an exclusively macrophage-restricted marker, and developmental recombination cannot be excluded. The authors should provide direct validation of reporter specificity in the adult meninges (e.g., co-labeling with established macrophage markers and exclusion of other Pf4-expressing lineages). At minimum, the limitations of Pf4Cre-based labeling should be discussed more explicitly, particularly regarding how off-target expression might affect Ca²⁺ signal interpretation.

The manuscript offers an extensive characterization of Ca²⁺ event features (frequency spectra, propagation patterns, synchrony), but the biological significance of these signals is largely speculative. There is no direct link established between Ca²⁺ activity patterns and macrophage function (e.g., activation state, motility, cytokine release, or interaction with other meningeal components). The discussion frequently implies functional specialization based on Ca²⁺ dynamics without experimental validation. To strengthen the conceptual impact, a clearer framing of the study as a foundational descriptive resource, rather than a functional dissection, would improve alignment between data and conclusions.

The GLM analysis revealing coupling between dural perivascular macrophage Ca²⁺ activity and vasomotion is technically sophisticated and intriguing. However, the directionality of this relationship remains unresolved. The current data do not distinguish whether macrophages actively regulate vasomotion, respond to mechanical or hemodynamic changes, or are co-modulated by neural activity. Statements suggesting that macrophages may "mediate" vasomotion are therefore premature. The authors should reframe these conclusions more cautiously, emphasizing correlation rather than causation, and expand the discussion to explicitly outline experimental strategies required to establish causality (e.g., macrophage-specific Ca²⁺ manipulation).

The authors conclude that synchronous Ca²⁺ events across macrophages are driven by extrinsic signals rather than intercellular communication, based primarily on distance-time analyses. This conclusion is not sufficiently supported, as spatial independence alone does not exclude paracrine signaling, vascular cues, or network-level coordination. No perturbation experiments are presented to test alternative mechanisms. The authors can either provide additional experimental evidence or rephrase the conclusion to acknowledge that the source of synchrony remains unresolved.

A major and potentially important finding is that the dominant macrophage response to CSD is a persistent decrease in Ca²⁺ activity, which is independent of CGRP/RAMP1 signaling. However, this phenomenon is not mechanistically explored. It remains unclear whether Ca²⁺ suppression reflects macrophage inhibition, altered viability, homeostatic resetting, or an anti-inflammatory program. Minimally, the discussion should be more deeply engaged with possible interpretations and implications of this finding.

The pharmacological blockade of RAMP1 supports a role for CGRP signaling in persistent Ca²⁺ increases after CSD, but the experiments are based on a relatively small number of cells and animals. The limited sample size constrains confidence in the generality of the conclusions. Pharmacological inhibition alone does not establish cell-autonomous effects in macrophages. The authors should acknowledge these limitations more explicitly and avoid overextension of the conclusions.

Reviewer #2 (Public review):

Using chronic intravital two-photon imaging of calcium dynamics in meningeal macrophages in Pf4Cre:TIGRE2.0-GCaMP6 mice, the study identified heterogeneous features of perivascular and non-perivascular meningeal macrophages at steady state and in response to cortical spreading depolarization (CSD). Analyses of calcium dynamics and blood vessels revealed a subpopulation of perivascular meningeal macrophages whose activity is coupled to behaviorally driven diameter fluctuations of their associated vessels. The analyses also investigated synchrony between different macrophage populations and revealed a role for CGRP/RAMP1 signaling in the CSD-induced increase, but not the decrease, in calcium transients.

This is a timely study at both the technical and conceptual levels, examining calcium dynamics of meningeal macrophages in vivo. The conclusions are well supported by the findings and will provide an important foundation for future research on immune cell dynamics within the meninges in vivo. The paper is well written and clearly presented.

I have only minor comments.

(1) Please indicate the formal definition of perivascular versus non-perivascular macrophages in terms of distance from the blood vessel. This information is not provided in the main text or the Methods. In addition, please explain how the meningeal vasculature was imaged in the main text.

(2) Similarly, the method used to induce acute CSD (pin prick) is not described in the main text and is only mentioned in the figure legends and Methods. Additional background on the neurobiology of acute CSD, as well as the resulting brain activity and neuroinflammatory responses, could be helpful.

Reviewer #3 (Public review):

Summary:

The authors of this report wish to show that distinct populations of meningeal macrophages respond to cortical spreading depolarization (CSD) via unique calcium activity patterns depending on their location in the meningeal sub-compartments. Perivascular macrophages display calcium signaling properties that are sometimes in opposition to non-perivascular macrophages. Many of the meningeal macrophages also displayed synchronous activity at variable distances from one another. Other macrophages were found to display calcium signals in response to dural vasomotion. CSD could induce variable calcium responses in both perivascular and non-perivascular macrophages in the meninges, in part due to RAMP1-dependent effects. Results will inform future research on the calcium responses displayed by macrophages in the meninges under both normal and pathological conditions.

Strengths:

Sophisticated in vivo imaging of meningeal immune cells is employed in the study, which has not been performed previously. A detailed analysis of the distinct calcium dynamics in various subtypes of meningeal macrophages is provided. Functional relevance of the responses is also noted in relation to CSD events.

Weaknesses:

The specificity of the methods used to target both meningeal macrophages and RAMP1 is limited. Additional discussion points on the functional relevance of the two subtypes of meningeal macrophages and their calcium responses are warranted. A section on potential pitfalls should be included.

Author Response:

The following is the authors’ response to the original reviews.

Public review:

Reviewer #1 (Public review):

Strengths:

(1) The use of chronic two-photon Ca2+ imaging in awake, behaving mice represents a major technical strength, minimizing confounds introduced by anesthesia. The development of a Pf4Cre:GCaMP6s reporter line, combined with high-resolution intravital imaging, enables long-term and subcellular analysis of macrophage Ca2+ dynamics in the meninges.

(2) The comparison between perivascular and non-perivascular macrophages reveals clear niche-dependent differences in Ca2+ signaling properties. The identification of macrophage Ca2+ activity temporally coupled to dural vasomotion is particularly intriguing and highlights a potential macrophage-vascular functional unit in the dura.

3) By linking macrophage Ca2+ responses to CSD and implicating CGRP/RAMP1 signaling in a subset of these responses, the study connects meningeal macrophage activity to clinically relevant neuroimmune pathways involved in migraine and other neurological disorders.

Thank you for recognizing the strengths in our work.

Weaknesses:

(1) The manuscript relies heavily on Pf4Cre-driven GCaMP6s expression to selectively image meningeal macrophages. Although prior studies are cited to support Pf4 specificity, Pf4 is not an exclusively macrophage-restricted marker, and developmental recombination cannot be excluded. The authors should provide direct validation of reporter specificity in the adult meninges (e.g., co-labeling with established macrophage markers and exclusion of other Pf4-expressing lineages). At minimum, the limitations of Pf4Cre-based labeling should be discussed more explicitly, particularly regarding how off-target expression might affect Ca2+ signal interpretation.

We acknowledge that PF4 is not an exclusively macrophage-restricted marker. Yet, among meningeal immunocytes, it is almost exclusively expressed in macrophages (1, 2). Furthermore, in the adult mouse meninges, PF4Cre-based reporter lines label nearly all dural and leptomeningeal macrophages and almost no other cells (3, 4). This Cre line has also been used to target border-associated macrophages (2, 4). Moreover, a recent study suggests that the bacterial artificial chromosome used to generate the PF4Cre line does not affect meningeal macrophage activity (4). Nonetheless, in the revised version, we discuss a potential limitation of the Pf4Cre-based labeling approach for studying meningeal macrophages’ Ca2+ signaling, namely that a very small population of other meningeal immune cells may also be labeled.

(2) The manuscript offers an extensive characterization of Ca2+ event features (frequency spectra, propagation patterns, synchrony), but the biological significance of these signals is largely speculative. There is no direct link established between Ca2+ activity patterns and macrophage function (e.g., activation state, motility, cytokine release, or interaction with other meningeal components). The discussion frequently implies functional specialization based on Ca2+ dynamics without experimental validation. To strengthen the conceptual impact, a clearer framing of the study as a foundational descriptive resource, rather than a functional dissection, would improve alignment between data and conclusions.

In our discussion, we indicated that “the exact link between the distinct Ca2+ signal properties of meningeal macrophage subsets observed herein and their homeostatic function remains to be established”. In the revised discussion part, we acknowledge that this is primarily a descriptive study that provides a foundational landscape of Ca2+ dynamics in meningeal macrophages.

(3) The GLM analysis revealing coupling between dural perivascular macrophage Ca2+ activity and vasomotion is technically sophisticated and intriguing. However, the directionality of this relationship remains unresolved. The current data do not distinguish whether macrophages actively regulate vasomotion, respond to mechanical or hemodynamic changes, or are co-modulated by neural activity. Statements suggesting that macrophages may "mediate" vasomotion are therefore premature. The authors should reframe these conclusions more cautiously, emphasizing correlation rather than causation, and expand the discussion to explicitly outline experimental strategies required to establish causality (e.g., macrophage-specific Ca2+ manipulation).

In the results section, we indicate that our data suggest that dural perivascular macrophages are functionally coupled to locomotion-driven dural vasomotion, either responding to it or mediating it. Furthermore, we discussed the possibilities that 1) macrophages sense vascular-related mechanical changes and 2) macrophage Ca2+ signaling regulates dural vasomotion. Moreover, we explicitly state that studying causality will require an experimental approach that has yet to be developed, enabling selective manipulation of dural perivascular macrophages.

(4) The authors conclude that synchronous Ca2+ events across macrophages are driven by extrinsic signals rather than intercellular communication, based primarily on distance-time analyses. This conclusion is not sufficiently supported, as spatial independence alone does not exclude paracrine signaling, vascular cues, or network-level coordination. No perturbation experiments are presented to test alternative mechanisms. The authors can either provide additional experimental evidence or rephrase the conclusion to acknowledge that the source of synchrony remains unresolved.

Thank you for this suggestion. In the revision, we indicate that further studies are required to resolve the exact source of synchrony.

(5) A major and potentially important finding is that the dominant macrophage response to CSD is a persistent decrease in Ca2+ activity, which is independent of CGRP/RAMP1 signaling. However, this phenomenon is not mechanistically explored. It remains unclear whether Ca2+ suppression reflects macrophage inhibition, altered viability, homeostatic resetting, or an anti-inflammatory program. Minimally, the discussion should be more deeply engaged with possible interpretations and implications of this finding.

While we propose that the decrease in macrophage Ca2+ signaling following CSD could indicate that a hyperexcitable cortex dampens meningeal immunity, in the revised discussion, we indicate that further studies are needed to determine whether this reduction in meningeal macrophage Ca2+ activity reflects altered viability or reduced immune function that could interfere with the macrophage’s ability to restore homeostasis and dampen local inflammation.

(6) The pharmacological blockade of RAMP1 supports a role for CGRP signaling in persistent Ca2+ increases after CSD, but the experiments are based on a relatively small number of cells and animals. The limited sample size constrains confidence in the generality of the conclusions. Pharmacological inhibition alone does not establish cell-autonomous effects in macrophages. The authors should acknowledge these limitations more explicitly and avoid overextension of the conclusions.

Although n=3 is common in intravital imaging of the meninges, including experiments employing pharmacological manipulations, such as RAMP1 inhibition (5-7), a larger sample size will increase confidence in the results. We further acknowledge that our pharmacological data indicate only a potential role for RAMP1 signaling in meningeal macrophages and that CGRP/RAMP1 signaling in other meningeal immune or vascular cells may also play a role.

Reviewer #2 (Public review):

Using chronic intravital two-photon imaging of calcium dynamics in meningeal macrophages in Pf4Cre:TIGRE2.0-GCaMP6 mice, the study identified heterogeneous features of perivascular and non-perivascular meningeal macrophages at steady state and in response to cortical spreading depolarization (CSD). Analyses of calcium dynamics and blood vessels revealed a subpopulation of perivascular meningeal macrophages whose activity is coupled to behaviorally driven diameter fluctuations of their associated vessels. The analyses also investigated synchrony between different macrophage populations and revealed a role for CGRP/RAMP1 signaling in the CSD-induced increase, but not the decrease, in calcium transients.

This is a timely study at both the technical and conceptual levels, examining calcium dynamics of meningeal macrophages in vivo. The conclusions are well supported by the findings and will provide an important foundation for future research on immune cell dynamics within the meninges in vivo. The paper is well written and clearly presented.

Thank you.

I have only minor comments.

(1) Please indicate the formal definition of perivascular versus non-perivascular macrophages in terms of distance from the blood vessel. This information is not provided in the main text or the Methods. In addition, please explain how the meningeal vasculature was imaged in the main text.

We did not measure the exact distance of the perivascular macrophages from the blood vessels, but defined them as such based on previous data showing that these cells reside along the abluminal surface and maintain tight interactions with mural cells (8). We now provide this information in the revised manuscript, including their labeling approach with a dextran tracer.

(2) Similarly, the method used to induce acute CSD (pin prick) is not described in the main text and is only mentioned in the figure legends and Methods. Additional background on the neurobiology of acute CSD, as well as the resulting brain activity and neuroinflammatory responses, could be helpful.

We have added more background and the method for inducing CSD (i.e., a pinprick in the frontal cortex) in the Results section.

Reviewer #3 (Public review):

Strengths:

Sophisticated in vivo imaging of meningeal immune cells is employed in the study, which has not been performed previously. A detailed analysis of the distinct calcium dynamics in various subtypes of meningeal macrophages is provided. Functional relevance of the responses is also noted in relation to CSD events.

Thank you for recognizing the strengths of our paper

Weaknesses:

(1) The specificity of the methods used to target both meningeal macrophages and RAMP1 is limited. Additional discussion points on the functional relevance of the two subtypes of meningeal macrophages and their calcium responses are warranted. A section on potential pitfalls should be included.

Please see previous responses regarding the specificity of the PF4Cre line for targeting macrophages. The specificity of the RAMP1 antagonist we used (BIBN4096, Olcegepant) has been confirmed by its developer Boehringer Ingelheim, and has been used to target CGRP signaling in numerous studies, including those targeting meningeal macrophage and vascular signaling (2, 7). A section on the study’s limitations has been added.

References:

(1) H. Van Hove et al., A single-cell atlas of mouse brain macrophages reveals unique transcriptional identities shaped by ontogeny and tissue environment. Nat Neurosci 22, 1021-1035 (2019).

(2) F. A. Pinho-Ribeiro et al., Bacteria hijack a meningeal neuroimmune axis to facilitate brain invasion. Nature 615, 472-481 (2023).

(3) G. L. McKinsey et al., A new genetic strategy for targeting microglia in development and disease. Elife 9, (2020).

(4) H. J. Barr et al., The circadian clock regulates scavenging of fluid-borne substrates by brain border-associated macrophages. bioRxiv, (2025).

(5) T. L. Roth et al., Transcranial amelioration of inflammation and cell death after brain injury. Nature 505, 223-228 (2014).

(6) M. V. Russo, L. L. Latour, D. B. McGavern, Distinct myeloid cell subsets promote meningeal remodeling and vascular repair after mild traumatic brain injury. Nat Immunol 19, 442-452 (2018).

(7) K. L. Monaghan et al., Highly dynamic dural sinuses support meningeal immunity. Nature, (2026).

(8) H. Min et al., Mural cells interact with macrophages in the dura mater to regulate CNS immune surveillance. J Exp Med 221, (2024).

Version 2

Abstract

The meninges, which envelop and protect the brain, host a dense network of resident macrophages with diverse roles in regulating homeostasis and neuroinflammation. Despite their importance, we have a limited understanding of their behavior in vivo. Many dynamic cellular functions of macrophages involve intracellular Ca2+ signaling. However, virtually nothing is known about the spatiotemporal Ca2+ dynamics of meningeal macrophages in vivo. We developed a chronic intravital two-photon imaging approach and related computational analysis tools to interrogate meningeal macrophage Ca2+ dynamics, at subcellular resolution, in a novel Pf4Cre:TIGRE2.0GCaMP6s/wt reporter mouse model. Using imaging in awake mice, we characterized Ca2+ activity in meningeal macrophages at steady state and in response to cortical spreading depolarization (CSD), an aberrant pro-inflammatory brain hyperexcitability event implicated in migraine, traumatic brain injury, and stroke. In homeostatic meninges, macrophages in the dural perivascular niche exhibited several Ca2+ dynamic features, including event duration and signal frequency spectrum, distinct from those of localized to the interstitial, non-perivascular niche. Simultaneous tracking of macrophage Ca2+ dynamics and local vasomotion revealed a subset of dural perivascular macrophages whose activity was coupled to locomotion-driven diameter fluctuations of their associated vessels. Most perivascular and non-perivascular meningeal macrophages displayed propagating intracellular Ca2+ activity and synchronized intercellular Ca2+ elevations, potentially driven by extrinsic factors. In response to CSD, the majority of perivascular and non-perivascular meningeal macrophages showed a persistent decrease in Ca2+ activity, while a smaller subset displayed Ca2+ elevations. Mechanistically, CGRP/RAMP1 signaling mediated the increase but not the decrease in CSD-mediated Ca2+ signaling. Collectively, our results highlight a previously unknown diversity of Ca2+ dynamics in meningeal macrophages at steady state and in response to an aberrant brain hyperexcitability event linked to neuroinflammation.

Introduction

Resident macrophages are key myeloid immunocytes that play an important role in innate immune surveillance and defense across various peripheral tissues and organs (1–3). The central nervous system also harbors a large subset of parenchymal macrophages, known as microglia, and several distinct subsets of macrophages localized to the brain’s border tissues, including the choroid plexus, perivascular spaces, and the meningeal compartments that cover, protect, and support the brain (4–8). Macrophages are the predominant immune cell type within the brain meninges, and recent studies have demonstrated their diverse ontogeny, transcriptomic profiles, and immune functions at steady state (5, 8–12) and in several neuropathological conditions (8, 13–16).

Cytoplasmic calcium (Ca2+) signaling underlies a wide variety of cellular homeostatic and inflammatory processes in macrophages (17–23). In addition to intracellular Ca2+ elevation, distinct spatiotemporal dynamics - including oscillation patterns, intracellular propagations, and intercellular synchronization of Ca2+ signals - may regulate different macrophage functions during steady state and pathophysiology (20–25). Despite our increased understanding of the diverse molecular signatures and contributions of meningeal macrophages to homeostasis and neuroinflammation, virtually nothing is known about their Ca2+ signaling heterogeneity in both healthy and diseased states.

Here, we comprehensively characterized the Ca2+ dynamics of individual macrophages localized to the brain meninges by combining intravital two-photon Ca2+ imaging in a novel reporter mouse line, in which the Ca2+ reporter GCaMP6s is expressed in platelet factor 4 (Pf4+) meningeal macrophages, with an event-based signaling analysis pipeline. Our data reveal several distinct spatiotemporal Ca2+ dynamic features in perivascular versus interstitial non-perivascular meningeal macrophages, including a unique coupling between the Ca2+ signals of dural perivascular macrophages and behaviorally-driven vasomotion of their associated dural vessels at steady state. Furthermore, our data uncover both increases and decreases in Ca2+ activity in distinct subsets of meningeal macrophages in response to cortical spreading depolarization (CSD), a pathophysiological brain hyperexcitability event linked to headache pain and neuroinflammation (26) in migraine, traumatic brain injury, and stroke (27). Mechanistically, our data suggest that calcitonin gene-related peptide/receptor activity-modifying protein 1 (CGRP/RAMP1) axis mediates CSD-evoked macrophage Ca2+ elevation and related brain-to-meninges neuroimmune signaling pathway, potentially involving CGRP released from sensitized meningeal sensory neurons acting on macrophage CGRP receptors (16, 28, 29).

Results

Characterizing macrophage Ca2+ signaling features in homeostatic brain meninges

Previous intravital imaging studies exploring the spatiotemporal dynamics of tissue-resident macrophages have primarily used CX3C motif chemokine receptor 1 (CX3CR1)-based mouse reporter strains (13, 22, 30, 31). However, other resident monocyte-derived cells are labeled in these reporter mice (32, 33). Moreover, brain microglia also express CX3CR1 (13, 33–35), limiting the use of these reporter mice for resolving the spatiotemporal subcellular Ca2+ dynamics of macrophages localized to the relatively thin meningeal layers covering the brain parenchyma. To systematically characterize meningeal macrophage Ca2+ dynamics in vivo and avoid contamination from Ca2+ signals arising from superficial parenchymal microglia, we leveraged recent findings showing that Pf4 is highly enriched in meningeal macrophages but not in other meningeal immunocytes or parenchymal microglia (10, 16, 36), and generated transgenic reporter mice expressing the highly sensitive Ca2+ indicator GCaMP6s (using TIGRE2.0-based Ai162D mice (37)) in Pf4+ macrophages (using Pf4Cre mice (38)). Notably, 100% of meningeal macrophages are labelled in Pf4Cre-based reporter mice (36).

Anesthetic agents impact intracellular Ca2+ signaling, including in brain macrophages and other non-excitable glial cells (39–41). We therefore investigated subcellular Ca2+ activity of meningeal macrophages in awake, behaving mice. We imaged meningeal macrophages via a chronic cranial window implanted together with a restraining headpost over the intact dura mater overlying the posterior neocortex. This chronic window approach produces minimal inflammatory responses in the cortex and meninges below the window (42, 43). After at least 7 days of recovery, mice were gradually habituated across multiple days to head restraint while free to run on a wheel (Figure 1b). We tracked Ca2+ transients of GCaMP6s-labeled meningeal macrophages using high-speed two-photon microscopy in 37 fields of view (FOVs) from 7 mice. We first corrected the imaging movies for locomotion-evoked meningeal translational shifts using rigid registration (43). Movies were then processed using the AQuA2 data analysis platform (44), which implements an unbiased event-based approach to capture spatiotemporal Ca2+ event dynamics (Figure 1C). Based on spatial analysis, we assigned events (n=1361) with their corresponding Ca2+ features to each macrophage, using data from 503 cells.

Imaging meningeal macrophage Ca2+ dynamics in awake behaving mice.

(A) Pf4Cre:GCaMP6sfl/wt reporter mouse construct for imaging meningeal macrophages Ca2+ activity.(B) Experimental procedure for two-photon imaging of meningeal macrophage Ca2+ activity. Following the implantation of a headpost and a cranial window, mice were habituated to head restraint and subjected to two-photon microscopy while head-fixed on a running wheel to study meningeal macrophage Ca2+ activity. (C) Macrophage Ca2+ imaging processing pipeline.

Meningeal macrophages occupy two distinct niches: perivascular (along the abluminal surface, physically contacting mural cells) and non-perivascular (within the interstitial space) (8, 45). These spatial distributions may dictate their divergent roles in meningeal immunity and vascular regulation. We therefore characterized the Ca2+ dynamics of these two anatomically distinct macrophage populations (perivascular, n=122; interstitial, non-perivascular, n=381, respectively, Figure 2A and Video S1). Most perivascular macrophages (93.4%, n=114) displaying ongoing Ca2+ activity were associated with vessels in the dura mater (labeled with a TRITC-Dextran tracer, see methods). The two meningeal macrophage subpopulations exhibited several distinct Ca2+ activity features. While the total area of Ca2+ activity in the peri- and non-perivascular macrophages was similar (Figure 2B), the signal perimeter of perivascular macrophages was significantly greater (Figure 2C) and exhibited a more elongated shape (Figure 2A, 2D), in agreement with their rod-shaped morphology (8, 46). The Ca2+ event duration in the perivascular macrophages was also longer compared to the interstitial macrophage subpopulation (Figure 2F). The peak Ca2+ activity level (Max ΔF/F0, Figure 2E) and event rate (Figure 2G) were, nonetheless, similar in the two meningeal macrophage subpopulations.

Ca2+ dynamic features of meningeal macrophages at steady state.

(A) Left: Mean projection of an example FOV depicting perivascular (P, red; 1, 2) and non-perivascular (NP, blue; 3, 4) meningeal macrophages (white squares). Scale bar 50 μm. Right: Corresponding macrophages with representative 900-second Ca2+ activity traces (top) and their fluorescence signal at selected time points (bottom). (B-G) AQuA2-based morphological and Ca2+ event functional features of perivascular (P=122 cells) and non-perivascular (NP, n=381 cells) meningeal macrophage, n=37 fields of view (FOVs) from 7 mice. (B) Event area, (C) Event perimeter, (D) Event circularity, (E) Event max DF/F0, (F) Event duration, (G) Event rate. Data (B-G) represent median ± IQR. ***p<0.001, ****p<0.0001, Mann-Whitney U-test.

Distinct Ca2+ signal frequency spectra may underlie different biological functions of macrophages (22). We therefore analyzed meningeal macrophage Ca2+ signal waveforms using a multi-step signal processing and clustering analysis. We clustered cells based on two distinct patterns of Ca2+ activity. Cells in Cluster 1 (n=40) exhibited a more noisy-like activity pattern characterized by multiple frequencies, while cells in Cluster 2 (n=463) displayed a single dominant frequency at 0.01 Hz (Figure 3A). To further explore these cellular signaling differences, we combined the clustering data with AQuA2-derived features and observed that macrophages in Cluster 1 showed a larger event perimeter, but lower circularity and peak magnitude than those in Cluster 2 (Figure 3D-F). While Cluster 1 cells had a lower signal-to-noise ratio compared to Cluster 2 cells (Figure 3I), both Clusters displayed similar event area, duration, and rate (Figure 3C, G, H). Finally, we observed a significant association between cell cluster (1 vs. 2) and cell type, with Cluster 1 predominantly comprising perivascular macrophages and Cluster 2 comprising primarily non-perivascular macrophages (Figure 3J), further suggesting that these two meningeal macrophage subpopulations have distinct Ca2+ signaling properties.

Intracellular Ca2+ signal frequency spectra of macrophage subsets in the steady-state meninges

(A) Left: Clustering of meningeal macrophage Ca2+ activity based on frequency-domain features and peak detection. Cluster 1 (purple, n=40 cells) and Cluster 2 (green, n=463 cells), n=37 fields of view (FOVs) from 7 mice. Right: Power spectrum density (PSD) of Ca2+ signals for each cluster. (B) Example DF/F0 heatmaps of Cluster 1 and Cluster 2 macrophages. (C-H) Morphological and Ca2+ functional features of Cluster 1 and Cluster 2 macrophages. (C) Event area. (D) Event perimeter (E) Event circularity. (F) Event max DF/F0. (G) Event duration. (H) Event rate. (I) Signal-to-noise ratio from Clusters 1 and 2. (J) Distribution of cell types across clusters. Data (C-I) represents median ± IQR. ***p<0.001, ****p < 0.0001, Mann-Whitney U-test Data (J) represents the cell proportion. ****p<0.0001, Chi-square test.

Intracellular Ca2+ signal propagation underlies diverse cellular functions and has been recently identified in macrophages in vitro (21) and in skin-resident macrophages in vivo (47). By assessing the Ca2+ signal propagation maps for each event within a defined cell, we identified two distinct patterns of activity: propagating events, in which Ca2+ signals traveled throughout the entire cell, and stationary events (Figure 4A). Propagating events showed varied signal source regions and directionality. Most macrophages (perivascular, 94.3%, n=115; non-perivascular, 86.9%, n=331) exhibited only propagating events, while a small minority of cells displayed a mix of propagating and stationary events or exclusively stationary activity events (Figure 4B).

Intracellular propagation and intercellular synchronization of meningeal macrophage Ca2+ activity.

(A) Spatial maps of two distinct Ca2+ events. Left: Propagating Ca2+ activity. Right: Stationary Ca2+ activity. (B) Distribution of event propagation profiles in perivascular (n=122 cells) and non-perivascular (n=381 cells) meningeal macrophages, n=37 fields of view (FOVs) from 7 mice. (C) Schematic analysis paradigm for detecting synchronous Ca2+ activity in meningeal macrophages. (D) Synchronous Ca2+ events among meningeal macrophages within a FOV. Left: Mean projection of an example FOV showing Ca2+ activity in distinct macrophages (colored/numbered). Scale bar 50 μm. Right: Spatial map of macrophages exhibiting synchronous Ca2+ activity. Lines connect macrophages with synchronized Ca2+ activity, and colors indicate the extent of Ca2+ event synchronization. (E) Distribution of distances across macrophage pairs showing different numbers of synchronous Ca2+ events. (F) Linear regression showing poor correlation between distances of macrophage with synchronized Ca2+ activity and event delay. (G) Proportion of macrophages exhibiting a specific synchronized interaction (cells interacting only with the same subtype: P-P/NP-NP; cells interacting only with a different subtype: P-NP; mixed (cells interacting with the same and different subtypes).

Macrophage intercellular communication, including synchronized activity, potentially across connected macrophage networks, has been implicated in maintaining tissue homeostasis and immune function (18, 21, 48, 49). To investigate meningeal macrophage intercellular interactions, we characterized spatiotemporal relationships between Ca2+ events in distinct cells within each FOV. We compared temporal factors, including the relative onset latency (t0A – t0B) between different macrophages and the duration of the first occurring event (t1A – t0A) (Figure 4C). We also compared the distances between macrophage pairs exhibiting concurrent events and the number of synchronous events (Figure 4D). Finally, we calculated the proportion of perivascular and non-perivascular macrophages exhibiting synchronous Ca2+ events. Across all FOVs, 49.3% of macrophages exhibited co-activation over 0–300 μm with minimal distance–delay correlation (Figure 4E, F), suggesting that spatial proximity does not influence event synchronicity. Both macrophage subtypes exhibit temporally coincident Ca2+ elevations (Figure 4G), consistent with a shared synchronization driver. The frequency of synchronous Ca2+ events detected could have been influenced by their duration (i.e., the longer the events, the higher the chance of detecting simultaneous event pairs). However, the duration of a given event was a poor predictor of the number of simultaneous events (Figure 4H).

Dural perivascular macrophage Ca2+ activity is tuned to behaviorally-driven dural vasomotion

Brain border-associated macrophages in the leptomeninges and related parenchymal perivascular spaces regulate pial arterial vasomotion indirectly by affecting vessel stiffness (6). Yet, interaction between vascular-associated macrophages residing in the dura mater, the outermost meningeal layer, and dural vasomotion remains unknown. We therefore imaged the dural vasculature (using a TRITC-Dextran tracer) together with macrophage Ca2+ activity and used a Generalized Linear Model (GLM) approach to investigate functional interaction between dural perivascular macrophage Ca2+ signals and dural vessel dynamics (Figure 5A). Dural arteries constrict during locomotion, while pial arteries dilate (50). We analyzed the locomotion-associated responses of 86 meningeal vessels (32 FOVs from 5 mice) and identified a subset (22%; n=19) in which the diameter changes were well fit by a GLM with locomotion state as a predictor. Of these, we identified 74% (n=14) as dural vessels based on their GLM’s negative coefficients consistent with constriction (Figure 5B and 5C). Next, we fitted the Ca2+ signal observed in perivascular macrophages associated with these dural vessels (n=35) to a GLM using the diameter changes as a predictor variable. Overall, the Ca2+ activity of 83% (n=29) of these dural macrophages was well predicted by the model (average deviance explained across all well-fit macrophages: 0.43±0.15, mean±SD). Analysis of the macrophage-vascular models’ beta coefficients revealed two distinct interactions. About half of the macrophages (55%, n=16) exhibited negative coefficients (i.e., increase and decrease in Ca2+ activity associated with dural vasoconstriction and recovery, respectively; Figure 5B and 5D). The remaining macrophages (45%, n=13) exhibited positive coefficients (i.e., decrease and increase in Ca2+ activity in response to dural constriction and recovery, respectively; Figure 5B and 5E). The coefficients for increased and decreased macrophage Ca2+ activity peaked near zero delay relative to the vasoconstriction and were not statistically different (Figure 5D-E and 5H). These data provide evidence that dural perivascular macrophages are functionally coupled to locomotion-driven dural vasomotion, either responding to or mediating it.

Ca2+ signals of dural perivascular macrophages are functionally coupled to behaviorally driven dural vasomotion.

(A) Experimental paradigm: Locomotion data were acquired during imaging in awake-behaving mice. Behaviorally-evoked changes in meningeal vessel diameter were obtained using segmentation of vessels labeled with a tracer during macrophage Ca2+ and further tested for coupling with Ca2+ signals. (B) Example data of macrophages with Ca2+ activity tuned to locomotion-related dural vessel vasomotion. Locomotion bouts (top trace) and dural vessel diameter (black trace) that was well fit by a GLM (red line) using locomotion state as a predictor. Note vasoconstriction during locomotion, indicating a dural vessel. The two bottom traces depict the Ca2+ signal (black traces) of distinct meningeal macrophages that were well-fit by a GLM using dural vessel diameter as a predictor, showing either a negative coefficient (purple) or a positive coefficient (green). (C) Temporal profile of dura vessels GLM coefficients averaged across all well-fitted vessels (n=19). Traces represent the median across all well-fit ROIs, with shaded regions indicating IQR. (D, E) Temporal profiles of GLM coefficient values for macrophages’ Ca2+ activity averaged across all well-fitted cells (purple, negative coefficient, n=16; green, positive coefficient, n=13). Traces represent the median across all well-fit macrophages, with shaded regions indicating IQR. (F) The deviance explained (goodness-of-fit estimates) of the vessel diameter data included in the GLM used to predict macrophage Ca2+ activity were not statistically different for macrophages showing negative (n=16) and positive (n=13) coefficients, suggesting similar interaction levels. (G) Comparison of peak positive and negative coefficients of well-fit macrophage Ca2+ activity/vasomotion GLMs. (H) Center of mass of GLM coefficients indicating that dural vessel diameter changes drive bidirectional changes in fluorescence signal at zero delay. Data (F-H) are median ± IQR. ****p<0.0001 (Mann-Whitney U-test). Data from 32 FOVs from 5 mice.

An acute aberrant pro-inflammatory brain hyperexcitability event drives diverse Ca2+ dynamics in meningeal macrophages

CSD is a slowly propagating depolarization of neurons and astrocytes that drastically disrupts transmembrane gradients and cortical synaptic activity. This aberrant brain hyperexcitability event has been linked to parenchymal inflammation and pain in migraine, traumatic brain injury, and stroke (26–28), and could also affect meningeal macrophages (51). In anesthetized mice subjected to a single CSD episode, a small subset of meningeal macrophages undergoes morphological changes resembling an inflammatory state (26). Given the direct anatomical and functional connections between the brain and meninges (11, 52) and the involvement of increased Ca2+ influx in macrophage inflammatory activation (53), we asked whether CSD drives intracellular Ca2+ elevations in meningeal macrophages. We used a pinprick stimulus in the frontal cortex to trigger a single CSD episode in awake mice (29, 54) and characterized the related changes in meningeal macrophage Ca2+ dynamics. In each experiment, we verified CSD induction based on the associated acute meningeal deformation and/or pial vasoconstriction observed in mice ((29) and Video S2). We studied CSD-related changes in Ca2+ dynamics in 249 macrophages (perivascular, n=64; non-perivascular, n=185; 13 FOVs from 10 mice). For each cell, we compared Ca2+ event rates during the passage of the CSD wave (1 min) and the PostCSD period (30 min) to baseline (PreCSD, 30 min) to assess acute and persistent Ca2+ responses, respectively. Given the low Ca2+ activity observed under steady state and the likelihood that no spontaneous Ca2+ elevations occur during the brief period of the CSD event, we considered cells to be either acutely activated or to exhibit an unchanged Ca2+ response (i.e., not activated). For studying more prolonged changes during the post-CSD period, we characterized cells as exhibiting persistently increased (event rate > 2× PreCSD), decreased (event rate < 0.5× PreCSD), or unchanged responses. These criteria were used to account for large, observable variations from baseline activity, while also minimizing the influence of spontaneous fluctuations observed in naïve mice. While consistent with previous studies on macrophages in different tissues (22), these changes were not intended to represent definitive biological criteria for Ca2+ activation and inhibition, but rather a descriptive categorization based on comparable individual cell data.

Using these criteria, we detected both acute and persistent Ca2+ activity changes following CSD (Figure 6 A-E and Video S2). While smaller subsets of meningeal macrophages exhibited acute (21.3%, n=53) and/or persistent increases (22.1%, n=55) in Ca2+ activity, we observed a persistent decrease in the majority of cells (58.6%, n=146). An acute increase was observed more often in peri-vascular macrophages (perivascular, 32.8%, n=21; non-perivascular, 17.3%, n=32, Figure 6F). Persistent changes in Ca2+ activity were similarly observed in peri- and non-perivascular macrophages (increases; perivascular, 28.2%, n=18; non-perivascular, 18.4%, n=37; decreases; perivascular, 50.0%, n=32; non-perivascular, 61.6%, n=114, Figure 6G). The macrophages’ propensity to develop a persistent Ca2+ increase was unrelated to their acute response (Figure 6H), suggesting that the mechanisms underlying these two temporal responses are distinct. However, cells that showed no acute activation were more likely to exhibit decreased Ca2+ activity post-CSD (Figure 6H). Finally, we observed that macrophages exhibiting a persistent increase in Ca2+ activity had lower baseline activity than those showing a persistent decrease (Figure 6I), suggesting that this post-CSD response is influenced by the macrophages’ basal Ca2+ activity.

Diverse meningeal macrophage Ca2+ dynamics following cortical spreading depression (CSD).

(A) Experimental setup and example data: (top) A small burr hole was drilled above the frontal cortex 7 days before Ca2+ imaging. Mice were pretreated with saline 30 min before imaging baseline macrophage Ca2+ activity (30 min, PreCSD). CSD was then induced with a pin prick, and macrophage Ca2+ activity was assessed during CSD (1 min during CSD), and post-CSD (30 min, post-CSD). (Bottom) A raster plot of macrophage Ca2+ activity showing the acute and persistent increases and persistent decrease in response to CSD. (B) Example of macrophage Ca2+ fluorescence changes following CSD. Images depict the mean projection over the specific experimental timeline. Arrows indicate an increase or a decrease in Ca2+ activity. Scale bar 50 μm. (C) Individual responses of perivascular (P, n=21) and non-perivascular (NP, n=32) macrophages showing an acute increase in Ca2+ activity. (D) Individual response of P (n=18) and NP (n=37) macrophages exhibiting a persistent increase in Ca2+ activity. (E) Individual responses of P (n=32) and NP (n=114) macrophages showing a persistent decrease in Ca2+ activity. (F) Proportion of P and NP macrophages showing an acute increase in Ca2+ activity or no acute change. (G) Proportion of P and NP macrophages showing a persistent increase, decrease, or no change in Ca2+ activity. (H) Proportion of macrophages displaying distinct persistent responses stratified based on their acute response. (I) Baseline (PreCSD) Ca2+ activity in macrophages exhibiting persistent increased (n=55) or decreased activity (n=146). Data (C-E) are median ± IQR. ****p<0.0001 (Wilcoxon signed rank test). Data (F-H) represents the proportion of cells, **p<0.01; ****p<0.0001 (Chi-square test). Data (I) are median ± IQR. ****p<0.0001 (Mann-Whitney U-test). CSD data from n=64 perivascular cells and n=185 non-perivascular cells; 13 FOVs from 10 mice.

CGRP/RAMP1 signaling mediates CSD-evoked persistent increase in meningeal macrophage Ca2+ activity

Many meningeal macrophages are localized near peptidergic, CGRP-expressing sensory axons (16). In the wake of CSD, cortex-to-meninges signaling enhances the responsiveness of meningeal sensory neurons that could drive CGRP release from their peripheral nerve endings (29, 54, 55). CGRP-expressing sensory neurons regulate tissue immunity and meningeal macrophage function via the CGRP/RAMP1 neuroimmune axis (16, 56). We therefore asked whether the CSD-related changes in meningeal macrophage Ca2+ dynamics we observed involve CGRP/RAMP1 signaling. We pretreated mice with the selective RAMP1 antagonist BIBN4096 and then imaged meningeal macrophage Ca2+ activity (42 cells; perivascular, n=14; non-perivascular, n=28; 3 FOVs from 3 mice) before and after CSD. As expected, RAMP1 blockade did not affect CSD triggering (57). Compared with the control saline treatment, RAMP1 blockade also did not reduce basal macrophage Ca2+ activity (Figure 7B). Blocking CGRP/RAMP1 signaling neither affected the incidence of acute increases in Ca2+ activity (Figure 7A and 7C) nor the magnitude of that response (Figure 7D). RAMP1 antagonism, however, inhibited the CSD-evoked persistent increase in the macrophage’s Ca2+ activity, without affecting the incidence of the persistent decrease (Figure 7E). The data suggest that in the wake of CSD, the CGRP/RAMP1 axis is responsible for the prolonged enhancement of Ca2+ signaling in a subset of meningeal macrophages, which could potentially mediate their pro-inflammatory response.

CGRP/RAMP1 signaling mediates CSD-related persistent increase in meningeal macrophage Ca2+ activity.

(A) Experimental setup and example data: (top) A small burr hole was drilled above the frontal cortex 7 days before Ca2+ imaging. Mice were pretreated with the RAMP1 antagonist BIBN4096 (BIBN) 30 min before imaging baseline macrophage Ca2+ activity (30 min, PreCSD). CSD was then induced with a pin prick, and macrophage Ca2+ activity was assessed during CSD (1 min during CSD), and post-CSD (30 min, post-CSD). (Bottom) A raster plot of macrophage Ca2+ activity showing the acute and persistent increases and persistent decrease in response to CSD. (B) RAMP1 inhibition does not affect the baseline (PreCSD) event rate. Data compared between macrophages imaged in saline-treated mice (n=249 cells, 13 FOVs from 10 mice) and BIBN-treated mice (n=42 cells, 3 FOVs from 3 mice). (C) RAMP1 antagonism does not affect the CSD-evoked acute increase in macrophage Ca2+. Proportion of macrophages showing an acute response (increase vs. no change) in saline- and BIBN-treated mice. (D) RAMP1 antagonism does not affect the magnitude of the acute macrophage Ca2+ signal. Event max DF/F0 in macrophages showing an acute Ca2+ increase in saline-treated mice (n=53 cells) and BIBN-treated mice (n=8 cells). (E) RAMP1 antagonism distinctly inhibits the persistent increase in macrophage Ca2+ activity post CSD. Proportion of macrophages showing persistent increase, persistent decrease, or no persistent change in saline-treated mice (n=249 cells) and BIBN-treated mice (n=42 cells). (F) Data (B, C, E) represents the proportion of cells. *p<0.05, Fisher’s exact test. Data (D) are median ± IQR.

Discussion

Resident macrophages in the brain meninges are essential for maintaining brain homeostasis, regulating central nervous system immune surveillance, and mediating neuroimmune responses under pathological conditions (12, 14, 16, 45, 58, 59). Macrophages rely on Ca2+ signaling to mediate many of their functions (17–23), yet remarkably little is known about the Ca2+ response properties of meningeal macrophages at steady state and disease. Here, using two-photon microscopy in awake behaving Pf4Cre:GCaMP6s reporter mice, we provide a foundational landscape of meningeal macrophage Ca2+ dynamics. We describe a heterogeneity of meningeal macrophage Ca2+ signals at steady state and in response to CSD, an aberrant cortical hyperexcitability event associated with migraine, traumatic brain injury, and stroke. Our data suggest that macrophages in discrete perivascular and interstitial non-perivascular meningeal niches exhibit several distinct Ca2+ signal properties at steady state. We further demonstrate Ca2+ activity in dural perivascular macrophages, which is tuned to behaviourally driven dural vasomotion. Finally, we describe opposing Ca2+ responses of meningeal macrophages following CSD, and demonstrate the contribution of the CGRP/RAMP1 axis in mediating CSD-evoked persistent Ca2+ elevations.

The exact link between the distinct Ca2+ signal properties of meningeal macrophage subsets observed herein and their homeostatic function remains to be established. The lower event magnitude and noisier signal observed in dural perivascular macrophages may reflect functional interactions with the pulsation dynamics of dural vessels (50). Indeed, by combining vascular and macrophage Ca2+ imaging, we demonstrate a tight temporal association between the diameter fluctuations of the dural vessels and the Ca2+ signal of their associated macrophages in awake, locomoting mice. Direct vascular-macrophage coupling may underlie this interaction, involving macrophages sensing vascular-related mechanical changes via Piezo1 signaling (60). Functional vascular-macrophage interaction may also involve mural cells as intermediators (45, 61). Studying whether macrophage Ca2+ signaling regulates dural vasomotion will require an experimental approach that has yet to be developed, enabling selective manipulation of perivascular dural macrophages. The paucity of ongoing Ca2+ activity in perivascular macrophages situated in the leptomeninges we observed supports recent findings that subdural perivascular macrophages indirectly affect pial and parenchymal vasomotion via extracellular matrix remodeling (6).

Intracellular Ca2+ signal propagation has been observed in various non-excitable cells, such as astrocytes (62). We show that the majority of meningeal macrophages, including perivascular and interstitial cells, exhibit intracellular Ca2+ signals that propagate throughout the entire cell, suggesting microdomain elevation of intracellular Ca2+ following release from internal stores. By characterizing the spatiotemporal relationships between Ca2+ signals in distinct cells, we also demonstrate synchronous events that are independent of spatial proximity, suggesting that synchronous Ca2+ activity is not driven by intercellular communication. Further studies will be required to resolve the exact source of synchrony. Interestingly, our data indicate that synchronized events involve both peri- and non-perivascular macrophages, despite distinct Ca2+ elevation patterns, suggesting that these meningeal macrophage subtypes similarly sense and respond to signals underlying synchronous activity.

Cortex-to-meninges signaling involves a relatively slow flow of soluble molecules within the cerebrospinal fluid that reach the subarachnoid space and then advance via arachnoid cuff exit points into the dura mater (11, 52). Our findings of acute Ca2+ elevation in a subset of extrasinusoidal perivascular dural macrophages coinciding with the CSD event suggest a rapid transfer of soluble signaling factors released from a hyperexcitable cortex across all meningeal layers (28). Nevertheless, we cannot exclude a mechanically driven macrophage response to the acute meningeal deformation produced by the neuronal and glial swelling and shrinkage of the cortical extracellular space during CSD (29, 63, 64). Our data also indicate a delayed, prolonged increase in Ca2+ signaling in a relatively small subset of macrophages post-CSD, which could underlie their proinflammatory-like morphological change (51, 53). Our findings also support the view that meningeal neuroimmune CGRP/RAMP1 axis serves as a mechanism responsible for this macrophage Ca2+ response, potentially via the activation of macrophage RAMP1/CLR receptor complex (16) by CGRP released from sensitized meningeal afferent axons (29, 54, 65). Whether the relatively small subset of meningeal macrophages featuring increased Ca2+ signaling serves a protective role (66, 67) or a proinflammatory, destructive function (68) remains to be elucidated. Intriguingly, our data points to a persistent decrease in macrophage Ca2+ activity post-CSD, not involving CGRP/RAMP1 signaling, as the most prevalent response. Further studies are needed to determine whether this reduction in Ca2+ activity reflects altered viability or reduced immune function that could interfere with the macrophage’s ability to restore homeostasis and dampen local inflammation (69).

There are several limitations to our study. First, while PF4Cre-based labeling has been shown to target brain border-associated macrophages, we cannot fully exclude the possibility that in a small subset of meningeal dendritic cells, monocytes, and T cells that have low-level PF4 expression (10), GCaMP6 was also expressed, leading to a Ca²⁺ signal. Nonetheless, a recent study using PF4CremTmG mice failed to detect EGFP reporter expression above background in any other meningeal cells by flow cytometry (70). Second, to enhance Ca2+ event detection, we downsampled the movies to ∼1 Hz.

We therefore could have missed fast Ca²⁺ transients or microdomain activity. Third, in our study, we imaged Ca2+ activity in extrasinusoidal meningeal macrophages. It is therefore possible that these cells exhibit distinct response properties compared to the subset of dural macrophages associated with the dural sinuses (8). Finally, our study used a pharmacological approach to determine whether CGRP/RAMP1 receptor signaling mediates macrophage Ca2+ responses to CSD. We acknowledge that this approach does not allow us to establish a specific role for macrophage CGRP signaling, given the possibility that CGRP/RAMP1 signaling in other meningeal vascular or immune cells (10, 16, 71) may indirectly affect the macrophage Ca2+ response.

Conclusions

We provide a detailed characterization of macrophage Ca2+ dynamics in homeostatic meninges, thereby expanding our understanding of their biological diversity. The coupling of dural perivascular macrophage Ca2+ signals and dural vasomotion may represent a unique homeostatic functional dural macrophage-vascular unit that controls dural perfusion. The diversity of meningeal macrophage Ca2+ responses to CSD further highlights the complexity of brain-to-meninges neuroimmune signaling and meningeal macrophage function in neurological disorders such as migraine, traumatic brain injury, and stroke. Our study also provides essential genetic and data analysis tools to further understand the molecular signaling underlying macrophage function at steady state and neuropathological conditions.

Methods

Animals

All experimental procedures were approved by the Beth Israel Deaconess Medical Center Institutional Animal Care and Use Committee. Experiments were conducted on adult Pf4Cre/wt:TIGRE2.0GCaMP6s/wt Ca2+ reporter mice (8-17 weeks) (9 males, 5 females). Mice were generated by crossing Pf4Cre (C57BL/6-Tg (Pf4-icre) Q3Rsko/J, Jackson laboratory, Strain #008535) mice with GCaMP6sfl/fl (Ai162D; B6.Cg-Igs7tm162.1(tetO-GCaMP6s,CAG-tTA2)Hze/J, Jackson laboratory, Strain #031562) mice. Animals were genotyped by Transnetyx Inc.

Surgical procedures

Animals were anesthetized using isoflurane in 100% O2 (induction: 3%; maintenance: 1.5-2%) and placed on a heating pad with a rectal probe attached to a stereotaxic frame to monitor animal body temperature during surgery. Animals received dexamethasone (8 mg/kg, i.p.) and Meloxicam SR (4 mg/kg, s.c.) to reduce inflammation and improve surgical outcomes. An eye ointment was used to prevent ocular drying. Mice were implanted with a titanium headpost and a 3 mm glass cranial window (1.5 mm lateral and 2 mm posterior to Bregma) over an intact dura covering the left posterior neocortex (29). Immediately after surgery, the mouse cage was placed on a water-circulating heating pad for faster recovery. Animals were then single-housed with access to a running wheel and a hut and allowed to recover for at least one week.

Wheel running acclimation

After the cranial window surgery, mice were allowed to recover for at least a week. To reduce stress associated with head-fixation during imaging and habituate to wheel running, the mice received multiple training sessions (10 min to 1 h over 3–4 days). In each session, the mouse was placed on a 3D printed running wheel, with its headpost attached to two clamps, and allowed to locomote freely.

Two-photon Ca2+ imaging

Awake-behaving mice were head-fixed to the running wheel by its headpost (Figure 1). We used a two-photon microscope (Neurolabware) with a Nikon 16X, 0.8 N.A. objective to acquire images at 15.5 Hz with digital zoom set at 4X (312 x 212 µm2 FOV). A MaiTai laser set to 920 nm with 25–40 mW power was used to excite fluorescence. The Scanbox package for MATLAB (Neurolabware) was used to control the microscope and acquire images and wheel running data. To image the meningeal vasculature, mice were administered 70 kDa TRITC-Dextran tracer (50 mg/kg, i.v.; Sigma-Aldrich).

Behavioral tracking during two-photon imaging

We recorded running speed in MATLAB using a custom-made encoder (Arduino) coupled to the 3D-printed running wheel. See below for details of analyses of behavioral variables.

Induction of CSD and pharmacological treatment

For CSD induction, a 1 mm burr hole was drilled at the frontal bone (1.5 mm anterior to the cranial window) to allow access to the brain cortical surface. A small amount of silicone elastomer (Kwik-Cast, WPI) was placed to cover the burr hole opening, and the animal was left to recover for at least one week. CSD was induced using a brief 2-second cortical pinprick (29). CSD induction was confirmed by the identification of a short-lasting meningeal deformation and/or transient pial constriction (29). On the experimental day, 30 minutes before baseline recording, mice were pretreated with the selective CGRP/RAMP1 antagonist BIBN4096 (0.3 ml, 1 mg/kg, i.p., Tocris) (55) or 0.3 ml of saline (Vehicle control).

Quantification and statistical analysis

Two-photon imaging movie processing

We used a discrete Fourier transform to perform rigid registration to correct for translation changes caused by brain motion during locomotion. Movies were then downsampled to 1.03 Hz. Locomotion signals were detected as described (43). All image processing and locomotion signal extraction were performed in MATLAB 2021b (Mathworks).

Ca2+ signal detection pipeline

For detecting macrophage Ca2+ signals, we used the Activity Quantification and Analysis (AQuA2) platform that implements an event-based approach with advanced machine learning techniques for temporal and spatial segmentation of Ca2+ fluorescence events 38. Importantly, this computational platform captures event dynamics beyond traditional ROI-based approaches. The following user-defined parameters were input: 0.49 µm/pixel spatial resolution, 1.03 Hz temporal resolution, 1-second minimal event duration detection; window of event size between 25% (125 µm2/ 523 pixels) and 150% (750 µm2/ 3140 pixels). Every event and cell identified was followed by a manual visual check. Subsequently, each cell was labelled as perivascular or non-perivascular according to its location relative to vessels. Morphological features (area, perimeter, circularity) and spatiotemporal aspects of the Ca2+ signals (i.e., ΔF/F0 dynamics, frequency, amplitude, duration) were used to analyze cell-specific characteristics and Ca2+ activity profiles. We employed the AQuA2 automatically-generated function (44) to characterize intracellular Ca2+ propagation. Intercellular Ca2+ activity was evaluated by analyzing temporally co-occurring events (synchronized event pairs), as well as their corresponding spatial localization in the FOV. All post-processing of AQuA2-generated data was performed using MATLAB 2021b (Mathworks).

Clustering of Ca2+ dynamics

We used a Savitzky-Golay filter to detrend and smooth Ca2+ activity traces. Polynomial order and frame size were optimized for each cell by selecting the combination of parameters that yielded a better signal-to-noise ratio. From the filtered signals, we extracted the dominant frequency using Fast Fourier Transform, and peak counts using minimum peak prominence (threshold of 10% signal amplitude range). The optimal number of clusters was determined using the Elbow method, after which k-means clustering was applied to group cells based on their signal characteristics.

Analysis of CSD-related changes in macrophage Ca2+ dynamics

For analyzing the effects of CSD on macrophages’ Ca2+ dynamics, we divided the activity of each cell into three phases: ‘PreCSD’ (minutes 0 – 30), ‘DuringCSD’ (minutes 30 – 31), and ‘PostCSD’ (minutes 31 – 61). Ca2+ event rates during CSD and PostCSD were compared with those of the PreCSD baseline to evaluate acute and persistent Ca2+ responses, respectively. Ca2+ responses were categorized as increased (event rate > 2x PreCSD), decreased (event rate < 0.5x PreCSD), or unchanged.

Analysis of locomotion

We extracted running speed (cm/sec) from the wheel encoder. To infer the locomotion state, we first concatenated all velocity signals obtained from a given mouse across all experiments and trained a two-state Hidden Markov Model using the MATLAB function ‘hmmtrain’. Then, the locomotion state was inferred for each individual imaging run by applying the MATLAB function ‘hmmviterbi’ with the model trained on the concatenated data. Locomotion bouts were defined as periods when the locomotion state was sustained for at least two seconds.

Vascular signals

We calculated changes in vascular diameter by first generating a maximum-intensity projection of the red channel and then drawing polygons (ROIs) around each vessel. Subsequently, for each frame, pixels inside each ROI were extracted, and a Radon transform was applied to get a 1D vessel profile. Radon transform of the first frame was used as a reference to normalize the results of subsequent frames, resulting in a time series of normalized diameter traces.

General linear models (GLM)

We investigate the functional interaction between dural perivascular macrophage Ca2+ activity and dural vasomotion by fitting a Gaussian GLMs using the GLMnet package (MATLAB 2021b), with elastic net regularization (α=0.01) and ten-fold cross-validation. We employed a two-step modeling. The first model was used to identify dural vessels by evaluating the correlation between changes in vessel diameter and locomotion state, and classifying dural or pial vessels based on their vasoconstriction and vasodilation dynamics, respectively (50). The second modeling step evaluated the relative contribution of perivascular dural Ca2+ activity signals to changes in dural vessel diameter. All signals were downsampled to 1.03 Hz to match the frame rate used when extracting fluorescence Ca2+ signals in AQuA2. To allow potential anticipatory or delayed responses of diameter-to-locomotion or Ca2+ fluorescence-to-diameter, the time interval used to analyze the response to the predictor was set from -60 s to +60 s. The GLM was trained on 75% of the data, and all predictions and model performance reported are from the remaining 25% testing set. A threshold of 0.1 goodness-of-fit deviance explained was set. For each predictor temporal shift, a response coefficient was generated. The centroid delay between the predictor and response was calculated, weighted by the absolute value of the coefficient (i.e., center-of-mass, COM).

Statistical analysis

All statistical analysis was performed using GraphPad Prism 10.4 and MATLAB 2021b. Data were analyzed using a Wilcoxon matched-pairs signed rank sum test or a Mann-Whitney U-test. Distribution of categorical data was analyzed using the Chi-square or Fisher’s exact test. P-values are indicated as follows: *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001.